Simple repression draft

### 2020-06-01 doc/section_04_bayesian_inference.tex
MRM pass through entire main text. Removed side notes. Still lots left

% !TEX root = ./bursty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference and model 
selection}
\label{section_04_bayesian_inference}

In this section of the paper, we continue our program of providing one complete
description of the entire broad sweep of studies that have been made in the
context of the repressor-operator model, dating all the way back to the original
work of Jacob and Monod and including the visionary quantitative work of
M\"{u}ller-Hill and collaborators~\cite{Oehler1990} and up to more recent
studies \cite{Garcia2011a}. In addition, the aim is to reconcile the equilibrium
and non-equilibrium perspectives that have been brought to bear on this problem.
From Section~\ref{section_02_means}, this reconciliation depends on a key
quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does
the free energy of repressor binding, as described in the equilibrium models and
indirectly inferred from gene expression measurements, agree with the
corresponding values of repressor binding and unbinding rates in the
non-equilibrium picture, measured or inferred more directly? In this section we
tackle the statistical inference problem of inferring these repressor rates from
mRNA FISH data. But before we can turn to the full case of simple repression, we
must choose an appropriate model of the constitutive promoter and infer the
parameter values in that model. This is the problem we address first.

@@

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
constitutive promoters, achieves the right balance of complexity and simplicity,
by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
problems of parameter degeneracy than model 4 in
Figure~\ref{fig2:constit_cartoons} suffered. Does this stand up to closer
scrutiny, namely, comparison to full mRNA distributions rather than simply their
moments? We will test this thoroughly on single-cell mRNA counts for different
unregulated promoters from Jones et.\ al.~\cite{Jones2014}.

@@

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
known as the \textit{conjugate} prior for a Poisson likelihood. This type of
priors have nice properties to work with, in particular a conjugate prior allows
us to get a closed form for the posterior distribution $p(\lambda \mid D)$. For
our case it can be shown that for a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution ~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_k m_k$ for notational convenience, and $N$ is the number of
cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

@@

We carried out MCMC sampling on the posterior of this model, starting with the
constitutive \textit{lacUV5} dataset from~\cite{Jones2014}. The resulting MCMC
samples are shown in Figure~\ref{fig:constit_post_full}(A). In contrast to the
active/inactive constitutive model considered in~\cite{Razo-Mejia2020}
(nonequilibrium model 4 in~\ref{fig2:constit_cartoons}), this model is
well-identified with both parameters determined to a fractional uncertainty of
5-10\%. The strong correlation reflects the fact that their product sets the
mean of the mRNA distribution, which is tightly constrained by the data, but
there is weak ``sloppiness''~\cite{Transtrum2015} along a set of values with a
similar product.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in orange. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. Single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
described in text. Each individual promoter is labeled with a unique number in
both (C) and (D) for cross comparison and for comparison with 
Figure~\ref{fig2:constit_cartoons}(B).}
\label{fig:constit_post_full}
\end{figure}

@@

\subsubsection{Connecting inferred parameters to prior work}
It is interesting to connect these speculations on $k_i$ and $b$ to the work
of~\cite{Brewster2012}, where these same 18 promoters were considered through
the lens of the three-state equilibrium model (model 2 in
Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
previously discussed the thermodynamic models of gene regulation can only make
statements about the mean gene expression. This implies that we can draw the
connection between both frameworks by equating the mean mRNA $\left\langle m
\right\rangle$. This results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma}
        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
\end{equation}
By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
\end{equation}
valid for all the binding energies considered here.

@@

\subsection{Transcription factor kinetics can be inferred from FISH
measurements}\label{sec:rep_kinetics_inference}
\subsubsection{Building the model and performing parameter inference}

Now that we have a satisfactory model in hand for constitutive promoters, we
would like to return to the main thread: can we reconcile the equilibrium and
nonequilibrium models by putting to the test
Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the correspondence between indirectly
inferred equilibrium binding energies and nonequilibrium kinetic rates? To make
this comparison, is it possible to infer repressor binding and unbinding rates
from mRNA distributions over a population of cells as measured by
single-molecule FISH in~\cite{Jones2014}? If so, how do these inferred rates
compare to direct single-molecule measurements such as from~\cite{Hammar2014}
and to binding energies such as from~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018}, which were inferred under the assumptions of the
equilibrium models in Figure~\ref{fig1:means_cartoons}(B)? And can this
comparison shed light on the unreasonable effectiveness of the equilibrium
models, for instance, in their application in~\cite{Chure2019, Chure2019a}?

@@

As we found in Section~\ref{sec:beyond_means}, for our purposes the ``right''
model of a constitutive promoter is the bursty picture, model five in
Figure~\ref{fig2:constit_cartoons}A. Therefore our starting point here is the
analogous model with repressor added, model 5 in
Figure~\ref{fig1:means_cartoons}(C). For a given repressor binding site and copy
number, this model has four rate parameters to be inferred: the repressor
binding and unbinding rates $k_R^+$, and $k_R^-$, the initiation rate of bursts,
$k_i$, and the mean burst size $b$ (we nondimensionalize all of these by the
mRNA degradation rate $\gamma$).

Before showing the mathematical formulation of our statistical inference model,
we would like to sketch the intuitive structure. The dataset
from~\cite{Jones2014} we consider consists of single-cell mRNA FISH measurements
of nine different conditions, spanning several combinations of three unique
repressor binding sites and four unique repressor copy numbers. We assume that
the values of $k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that these values
are the same across datasets with different repressor binding sites and copy
numbers. In other words, we assume that the regulation of the transcription
factor does not affect the mean burst size nor the burst initiation rate. The
regulation happens as the promoter is taken away from the transcriptionally
active state when the promoter is bound by repressor. We assume that there is one unbinding
rate parameter for each repressor binding site, which is shared across aTc
inducer conditions, and likewise one binding rate for each unique repressor copy
number. This makes our model seven dimensional, or nine if one counts $k_i$ and
$b$ as well. Note that we use only a subset of the datasets from
Jones et.\ al.~\cite{Jones2014}, as discussed more in Appendix~\ref{sec:bayes}.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\},
\end{equation}
where subscripts for dissociation rates $k^-$  indicate the different repressor
binding sites, and subscripts for association rates $k^+$ indicate the aTc
concentration at which the repressor itself was induced. We emphasize that for
this particular dataset the repressor copy numbers are not known directly, so we
label their association rates by the concentration of a small inducer molecule 
aTc which indirectly controls the transcription of $\textit{lacI}$, as
described in detail in~\cite{Jones2014}. Also note
that the authors of~\cite{Jones2014} report estimates of LacI copy number per
cell rather than direct measurements. However, these estimates were made
\textit{assuming} the validity of the equilibrium models in
Figure~\ref{fig1:means_cartoons}, and since \textit{testing} these models is our
present goal, it would be circular logic if we were to make the same assumption.
Therefore we will make no assumptions about the LacI copy number for given aTc
concentrations.

@@

\mmnote{Probably SI most of this paragraph?}
This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
we instead resorted to numerics. One insight we found
\mmnote{a plot would probably help explain, maybe SI, need to explain better}
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
values, except with $k_i$ multiplied by the fold-change
$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the
\textit{ratio} $k_R^+/k_R^-$ was detectable. But for intermediate repression,
the distribution was visibly broadened with Fano factor greater than $1+b$, the
value for the corresponding constitutive case. This indicates that the repressor
rates had left an imprint on the distribution, and perhaps intuitively, this
intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
$k_i$. Put another way, if the repressor rates are much faster or much slower
than $k_i$, then there is a timescale separation and effectively only one
timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
rates in the problem are comparable does the mRNA distribution retain detectable
information about them.

@@

Figure~\ref{fig4:repressed_post_full}(B) shows both comparisons, with the top
panel comparing to equilibrium binding energies and the bottom panel comparing
to single-molecule measurements. First consider the top panel and its comparison
between repressor kinetic rates and binding energies. As described in
section~\ref{section_02_means}, if the equilibrium binding energies
from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018} indeed are the physically real
binding energies we believe them to be, then they should be related to the
repressor kinetic rates via Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, which we
restate here,
\begin{equation}
\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
        = - \log(k_R^+/k_R^-).
\label{eq:deltaFR_eq_noneq_equiv_repeat}
\end{equation}
Assuming mass action kinetics implies that $k_R^+$ is proportional to repressor
copy number $R$, or more precisely, it can be thought of as repressor copy
number times some intrinsic per molecule association rate. But since $R$ is not
directly known for our data from~\cite{Jones2014}, we cannot use this equation
directly. Instead we can consider two different repressor binding sites and
compute the \textit{difference} in binding energy between them, since this
difference depends only on the unbinding rates and not on the binding rates.
This can be seen by evaluating Eq.~\ref{eq:deltaFR_eq_noneq_equiv_repeat} for
two different repressor binding sites, labeled (1) and (2), but with the same
repressor copy number $R$, and taking the difference to find
\begin{equation}
\Delta F_R^{(1)} - \Delta F_R^{(2)}
= \beta\Delta\varepsilon_1 - \beta\Delta\varepsilon_2
= - \log(k_R^+/k_1^-) + \log(k_R^+/k_2^-),
\end{equation}
or simply
\begin{equation}
\beta\Delta\varepsilon_1 - \beta\Delta\varepsilon_2
= \log(k_2^-/k_1^-).
\end{equation}
The left and right hand sides of this equation are exactly the horizontal and
vertical axes of the top panel of Figure~\ref{fig4:repressed_post_full}. Since
we inferred rates for three repressor binding sites (O1, O2, and Oid), there are
only two independent differences that can be constructed, and we arbitrarily
chose to plot O2-O1 and O1-Oid in Figure~\ref{fig4:repressed_post_full}(B).
Numerically, we compute values of $k_{O1}^- / k_{Oid}^-$ and $k_{O2}^- /
k_{O1}^-$ directly from our full posterior samples, which conveniently provides
uncertainties as well, as detailed in Appendix~\ref{sec:bayesian}.
\mmnote{Explain in appendix how marginalization with sampling makes this
``trivial'' to compute from samples of the joint posterior, b/c sampling makes
marginalization trivial.} We then compare these log ratios of rates to the
binding energy differences $\Delta\varepsilon_{O1} - \Delta\varepsilon_{Oid}$
and from $\Delta\varepsilon_{O2} - \Delta\varepsilon_{O1}$ as computed from the
values from both~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}. Three of the four
values are within $\sim0.5 k_BT$ of the diagonal representing perfect agreement,
which is comparable to the $\sim\text{few}\times0.1 k_BT$ variability between
the independent determinations of the same quantities between~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018}. The only outlier involves Oid measurements
from~\cite{Razo-Mejia2018}, and as the authors of~\cite{Razo-Mejia2018} note,
this is a difficult measurement of low fluorescence signal against high
background since Oid represses so strongly. We are therefore inclined to regard
the failure of this point to fall near the diagonal as a testament to the
difficulty of the measurement and not as a failure of our theory. \mmnote{Does
this seem a fair characterization?   - RP: thanks, that was very helpful for
me.}

@@

\subsubsection{Comparison with prior measurements of repressor kinetics}
In the previous section we established the equivalence between the equilibrium
models' binding energies and the repressor kinetics we infer from mRNA
population distributions. But one might worry that the repressor rates we infer
from mRNA distributions are \textit{themselves} merely fitting parameters and
that they do not actually correspond to the binding and unbinding rates of the
repressor in vivo. To verify that this is not the case, we next compare our
kinetic rates with a different measurement of the same rates using a radically
different method: single molecule measurements as performed in Hammar et.\
al.~\cite{Hammar2014}. This is plotted in the lower panel of
Figure~\ref{fig4:repressed_post_full}B.

@@

The degree of quantitative agreement in the lower panel of
Figure~\ref{fig4:repressed_post_full}B clearly depends on the precise choice of
$\gamma$. Nevertheless we find this comparison very satisfying, when two wildly
different approaches to a measurement of the same quantity yield broadly
compatible results. We emphasize the agreement between our rates and the rates
reported in~\cite{Hammar2014} for any reasonable $\gamma$: values differ by at
most a factor of 2 and possibly agree to within our uncertainties of 10-20\%.
From this we feel confident asserting that the parameters we have inferred from
Jones et.\ al.'s single-cell mRNA counts data do in fact correspond to repressor
binding and unbinding rates, and therefore our conclusions on the agreement of
these rates with binding energies from~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018} are valid.

\subsubsection{Model checking}
In Figure~\ref{fig:constit_post_full}B we saw that the simple Poisson model of a
constitutive promoter, despite having a well behaved posterior, was clearly
insufficient to describe the data. It behooves us to carry out a similar check
for our model of simple repression, codified by Eq.~\ref{eq:p_m_bursty+rep} for
the steady-state mRNA copy number distribution. As derived in
Sections~\ref{section_02_means} and~\ref{sec:beyond_means}, we have compelling
theoretical reasons to believe it is a good model, but if it nevertheless turned
out to be badly contradicted by the data we should like to know.

>>>>

% !TEX root = ./bursty_transcription.tex
\section{Finding the ``right'' model: Bayesian parameter inference}
\label{section_04_bayesian_inference}

In this section of the paper, we continue our program of providing one complete
description of the entire broad sweep of studies that have been made in the
context of the repressor-operator model, dating all the way back to the original
work of Jacob and Monod and including the visionary quantitative work of
M\"{u}ller-Hill and collaborators~\cite{Oehler1990} and up to more recent
studies \cite{Garcia2011a}. In addition, the aim is to reconcile the equilibrium
and non-equilibrium perspectives that have been brought to bear on this problem.
From Section~\ref{section_02_means}, this reconciliation depends on a key
quantitative question as codified by Eq.~\ref{eq:deltaFR_eq_noneq_equiv}: does
the free energy of repressor binding, as described in the equilibrium models and
indirectly inferred from gene expression measurements, agree with the
corresponding values of repressor binding and unbinding rates in the
non-equilibrium picture, measured or inferred more directly? In this section we
tackle the statistical inference problem of inferring these repressor rates from
single-cell mRNA counts data. But before we can turn to the full case of simple
repression, we must choose an appropriate model of the constitutive promoter and
infer the parameter values in that model. This is the problem we address first.

@@

From consideration of Fano factors in the previous section, we suspect that
model 5 in Figure~\ref{fig2:constit_cartoons}(A), a one-state bursty model of
constitutive promoters, achieves the right balance of complexity and simplicity,
by allowing both Fano factor $\nu>1$, but also by remedying, by design, the
problems of parameter degeneracy than model 4 in
Figure~\ref{fig2:constit_cartoons} suffered~\cite{Razo-Mejia2020}. Does this
stand up to closer scrutiny, namely, comparison to full mRNA distributions
rather than simply their moments? We will test this thoroughly on single-cell
mRNA counts for different unregulated promoters from Jones et.\
al.~\cite{Jones2014}.

@@

As detailed in Appendix~\ref{sec:bayesian} this particular choice of prior is
known as the \textit{conjugate} prior for a Poisson likelihood. This type of
priors have nice properties to work with, in particular a conjugate prior allows
us to get a closed form for the posterior distribution $p(\lambda \mid D)$. For
our case it can be shown that for a Poisson distribution likelihood with its
Gamma distribution conjugate prior, the posterior distribution is also a Gamma
distribution ~\cite{Gelman2013}. Specifically the two parameters $\alpha'$ and
$\beta'$ for this posterior distribution take the form $\alpha' = \alpha +
\bar{m} N$ and $\beta' = \beta + N$, where we defined the sample mean $\bar{m} =
\frac{1}{N}\sum_{k=1}^N m_k$ for notational convenience, and $N$ is the number
of cells in our dataset. Furthermore, given that $N$ is $\mathcal{O}(10^3)$ and
$\langle m\rangle \gtrsim 0.1$ for all promoters measured in~\cite{Jones2014}
our data easily overwhelms the choice of prior, and allows us to approximate the
Gamma distribution with a Gaussian distribution with mean $\bar{m}$ and variance
$\bar{m} / N$ with marginal errors. As an example with real numbers, for the
\textit{lacUV5} promoter, Jones et.\ al~\cite{Jones2014} measured 2648 cells
with an average mRNA count per cell of $\bar{m} \approx 18.7$. For this case our
posterior distribution $P(\lambda \mid D)$ would be a Gaussian distribution with
mean $\mu = 18.7$, and a standard deviation $\sigma \approx 0.08$. This suggests
we have inferred our model's one parameter to a precision of order 1\%.

@@

We carried out Markov-Chain Monte Carlo (MCMC) sampling on the posterior of this
model, starting with the constitutive \textit{lacUV5} dataset
from~\cite{Jones2014}. The resulting MCMC samples are shown in
Figure~\ref{fig:constit_post_full}(A). In contrast to the active/inactive
constitutive model considered in~\cite{Razo-Mejia2020} (nonequilibrium model 4
in~\ref{fig2:constit_cartoons}), this model is well-identified with both
parameters determined to a fractional uncertainty of 5-10\%. The strong
correlation reflects the fact that their product sets the mean of the mRNA
distribution, which is tightly constrained by the data, but there is weak
``sloppiness''~\cite{Transtrum2015} along a set of values with a similar
product.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in black. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. Single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
described in text. Each individual promoter is labeled with a unique number in
both (C) and (D) for cross comparison and for comparison with 
Figure~\ref{fig2:constit_cartoons}(B).}
\label{fig:constit_post_full}
\end{figure}

@@

\subsubsection{Connecting inferred parameters to prior work}
It is interesting to connect these inferences on $k_i$ and $b$ to the work
of~\cite{Brewster2012}, where these same 18 promoters were considered through
the lens of the three-state equilibrium model (model 2 in
Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
previously discussed the thermodynamic models of gene regulation can only make
statements about the mean gene expression. This implies that we can draw the
connection between both frameworks by equating the mean mRNA $\left\langle m
\right\rangle$. This results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma}
        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
\end{equation}
By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
\end{equation}
valid for all the binding energies considered here.

@@

\subsection{Transcription factor kinetics can be inferred from single-cell mRNA
distribution measurements}\label{sec:rep_kinetics_inference}
\subsubsection{Building the model and performing parameter inference}

Now that we have a satisfactory model in hand for constitutive promoters, we
would like to return to the main thread: can we reconcile the equilibrium and
nonequilibrium models by putting to the test
Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the correspondence between indirectly
inferred equilibrium binding energies and nonequilibrium kinetic rates? To make
this comparison, is it possible to infer repressor binding and unbinding rates
from mRNA distributions over a population of cells as measured by
single-molecule Fluorescence \textit{in situ} Hybridization in~\cite{Jones2014}?
If so, how do these inferred rates compare to direct single-molecule
measurements such as from~\cite{Hammar2014} and to binding energies such as
from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}, which were inferred under the
assumptions of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
And can this comparison shed light on the unreasonable effectiveness of the
equilibrium models, for instance, in their application in~\cite{Chure2019,
Chure2019a}?

@@

As we found in Section~\ref{sec:beyond_means}, for our purposes the ``right''
model of a constitutive promoter is the bursty picture, model five in
Figure~\ref{fig2:constit_cartoons}(A). Therefore our starting point here is the
analogous model with repressor added, model 5 in
Figure~\ref{fig1:means_cartoons}(C). For a given repressor binding site and copy
number, this model has four rate parameters to be inferred: the repressor
binding and unbinding rates $k_R^+$, and $k_R^-$, the initiation rate of bursts,
$k_i$, and the mean burst size $b$ (we nondimensionalize all of these by the
mRNA degradation rate $\gamma$).

Before showing the mathematical formulation of our statistical inference model,
we would like to sketch the intuitive structure. The dataset
from~\cite{Jones2014} we consider consists of single-cell mRNA counts data of
nine different conditions, spanning several combinations of three unique
repressor binding sites and four unique repressor copy numbers. We assume that
the values of $k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that these values
are the same across datasets with different repressor binding sites and copy
numbers. In other words, we assume that the regulation of the transcription
factor does not affect the mean burst size nor the burst initiation rate. The
regulation happens as the promoter is taken away from the transcriptionally
active state when the promoter is bound by repressor. We assume that there is
one unbinding rate parameter for each repressor binding site, and likewise one
binding rate for each unique repressor copy number. This makes our model seven
dimensional, or nine if one counts $k_i$ and $b$ as well. Note that we use only
a subset of the datasets from Jones et.\ al.~\cite{Jones2014}, as discussed more
in Appendix~\ref{sec:bayes}.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\},
\end{equation}
where subscripts for dissociation rates $k^-$  indicate the different repressor
binding sites, and subscripts for association rates $k^+$ indicate the
concentration of the small-molecule that controlled the expression of the LacI
repressor (see Appendix~\ref{sec:bayes}). This is because for this particular
dataset the repressor copy numbers were not measured directly, but it is safe to
assume that a given concentration of the inducer resulted in a specific mean
repressor copy number\cite{Chure2019a}. Also note that the authors
of~\cite{Jones2014} report estimates of LacI copy number per cell rather than
direct measurements. However, these estimates were made assuming the validity of
the equilibrium models in Figure~\ref{fig1:means_cartoons}, and since testing
these models is our present goal, it would be circular logic if we were to make
the same assumption. Therefore we will make no assumptions about the LacI copy
number for a given inducer concentrations.

@@

This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
we instead resorted to numerics. One insight we found
\mmnote{a plot would probably help explain, maybe SI, need to explain better}
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
values, except with $k_i$ multiplied by the fold-change
$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the ratio
$k_R^+/k_R^-$ was detectable. But for intermediate repression, the distribution
was visibly broadened with Fano factor greater than $1+b$, the value for the
corresponding constitutive case. This indicates that the repressor rates had
left an imprint on the distribution, and perhaps intuitively, this intermediate
regime occurs for values of $k_R^\pm$ comparable to the burst rate $k_i$. Put
another way, if the repressor rates are much faster or much slower than $k_i$,
then there is a timescale separation and effectively only one timescale remains,
$k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three rates in the problem
are comparable does the mRNA distribution retain detectable information about
them.

@@

Figure~\ref{fig4:repressed_post_full}(B) shows both comparisons, with the top
panel comparing to equilibrium binding energies and the bottom panel comparing
to single-molecule measurements. First consider the top panel and its comparison
between repressor kinetic rates and binding energies. As described in
section~\ref{section_02_means}, if the equilibrium binding energies
from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018} indeed are the physically real
binding energies we believe them to be, then they should be related to the
repressor kinetic rates via Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, which we
restate here,
\begin{equation}
\Delta F_R = \beta\Delta\varepsilon_R - \log(R/N_{NS})
        = - \log(k_R^+/k_R^-).
\label{eq:deltaFR_eq_noneq_equiv_repeat}
\end{equation}
Assuming mass action kinetics implies that $k_R^+$ is proportional to repressor
copy number $R$, or more precisely, it can be thought of as repressor copy
number times some intrinsic per molecule association rate. But since $R$ is not
directly known for our data from~\cite{Jones2014}, we cannot use this equation
directly. Instead we can consider two different repressor binding sites and
compute the \textit{difference} in binding energy between them, since this
difference depends only on the unbinding rates and not on the binding rates.
This can be seen by evaluating Eq.~\ref{eq:deltaFR_eq_noneq_equiv_repeat} for
two different repressor binding sites, labeled (1) and (2), but with the same
repressor copy number $R$, and taking the difference to find
\begin{equation}
\Delta F_R^{(1)} - \Delta F_R^{(2)}
= \beta\Delta\varepsilon_1 - \beta\Delta\varepsilon_2
= - \log(k_R^+/k_1^-) + \log(k_R^+/k_2^-),
\end{equation}
or simply
\begin{equation}
\beta\Delta\varepsilon_1 - \beta\Delta\varepsilon_2
= \log(k_2^-/k_1^-).
\end{equation}
The left and right hand sides of this equation are exactly the horizontal and
vertical axes of the top panel of Figure~\ref{fig4:repressed_post_full}. Since
we inferred rates for three repressor binding sites (O1, O2, and Oid), there are
only two independent differences that can be constructed, and we arbitrarily
chose to plot O2-O1 and O1-Oid in Figure~\ref{fig4:repressed_post_full}(B).
Numerically, we compute values of $k_{O1}^- / k_{Oid}^-$ and $k_{O2}^- /
k_{O1}^-$ directly from our full posterior samples, which conveniently provides
uncertainties as well, as detailed in Appendix~\ref{sec:bayesian}.
\mmnote{Explain in appendix how marginalization with sampling makes this
``trivial'' to compute from samples of the joint posterior, b/c sampling makes
marginalization trivial.} We then compare these log ratios of rates to the
binding energy differences $\Delta\varepsilon_{O1} - \Delta\varepsilon_{Oid}$
and from $\Delta\varepsilon_{O2} - \Delta\varepsilon_{O1}$ as computed from the
values from both~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}. Three of the four
values are within $\sim0.5 k_BT$ of the diagonal representing perfect agreement,
which is comparable to the $\sim\text{few}\times0.1 k_BT$ variability between
the independent determinations of the same quantities between~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018}. The only outlier involves Oid measurements
from~\cite{Razo-Mejia2018}, and as the authors of~\cite{Razo-Mejia2018} note,
this is a difficult measurement of low fluorescence signal against high
background since Oid represses so strongly. We are therefore inclined to regard
the failure of this point to fall near the diagonal as a testament to the
difficulty of the measurement and not as a failure of our theory.

@@

\subsubsection{Comparison with prior measurements of repressor kinetics}
In the previous section we established the equivalence between the equilibrium
models' binding energies and the repressor kinetics we infer from mRNA
population distributions. But one might worry that the repressor rates we infer
from mRNA distributions are \textit{themselves} merely fitting parameters and
that they do not actually correspond to the binding and unbinding rates of the
repressor in vivo. To verify that this is not the case, we next compare our
kinetic rates with a different measurement of the same rates using a radically
different method: single molecule measurements as performed in Hammar et.\
al.~\cite{Hammar2014}. This is plotted in the lower panel of
Figure~\ref{fig4:repressed_post_full}(B).

@@

The degree of quantitative agreement in the lower panel of
Figure~\ref{fig4:repressed_post_full}(B) clearly depends on the precise choice
of $\gamma$. Nevertheless we find this comparison very satisfying, when two
wildly different approaches to a measurement of the same quantity yield broadly
compatible results. We emphasize the agreement between our rates and the rates
reported in~\cite{Hammar2014} for any reasonable $\gamma$: values differ by at
most a factor of 2 and possibly agree to within our uncertainties of 10-20\%.
From this we feel confident asserting that the parameters we have inferred from
Jones et.\ al.'s single-cell mRNA counts data do in fact correspond to repressor
binding and unbinding rates, and therefore our conclusions on the agreement of
these rates with binding energies from~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018} are valid.

\subsubsection{Model checking}
In Figure~\ref{fig:constit_post_full}(B) we saw that the simple Poisson model of
a constitutive promoter, despite having a well behaved posterior, was clearly
insufficient to describe the data. It behooves us to carry out a similar check
for our model of simple repression, codified by Eq.~\ref{eq:p_m_bursty+rep} for
the steady-state mRNA copy number distribution. As derived in
Sections~\ref{section_02_means} and~\ref{sec:beyond_means}, we have compelling
theoretical reasons to believe it is a good model, but if it nevertheless turned
out to be badly contradicted by the data we should like to know.
